Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-04 talk/dls2014/paper/paper.tex
clarification
Add a comment
first round of feedback
minor changes before restructuring

We focus on the \emph{threading} approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} are implementations of this. It requires
great care in order to avoid deadlocks, which is why we follow the TM
approach that provides a \emph{direct} replacement for the GIL. It
does not require careful placing of locks in the right spots. We will
compare our work with Jython for evaluation.

@@

The x86 CPUs provide a feature which is precisely called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment register
$\%gs$, which is mostly unused in current applications, to point to the
thread's segment start address.\footnote{The other segment register
$\%fs$ is typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for a
similar purpose: thread-local data -- but a lot of it.}  The translation
from a $SO$ to a LA is done for us by using the ``$\%gs\colon$'' prefix
before any CPU instruction that reads or writes memory using a $SO$
address.  This is very efficient and some compilers support it natively
(e.g.\ clang).

@@

Conflicts can be resolved by aborting one of the two transactions, or in
some cases by pausing one of them until the other commits.  If a commit
succeeds, we copy the changes to modified objects so that they become
visible in other segments.  Note that this only includes objects that
existed before the transaction and have been modified by it.  It does
not includes all the new objects created by the current transaction
(there is typically a very high rate of object churn in dynamic languages).
Our garbage collector allocates new objects in pages that are shared in
the first place, so that they do not need to be copied -- and moreover,
such objects will commonly never be modified any more.

@@

\begin{lstlisting}
void stm_start_transaction(tl, jmpbuf)
void stm_commit_transaction()
void stm_read(object_t *obj)
void stm_write(object_t *obj)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(tl, obj)
STM_POP_ROOT(tl, obj)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction. It requires
two arguments, the first being a thread-local data structure and the
second a buffer for use by \lstinline!setjmp()!.
\lstinline!stm_commit_transaction()!  tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on an object and \lstinline!stm_allocate()!
allocates a new object with the specified size.
 \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
and pop objects on the shadow stack.\footnote{A stack for pointers to
  GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}}  Objects have to be saved using this stack around calls
that may cause a GC cycle to happen, and also while there is no
transaction running. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

The type \lstinline!object_t!  is special as it causes the
compiler\footnote{We use Clang 3.5 with some patches to fix bugs in its address-space
 256 feature.} to make all accesses through it relative to the $\%gs$
register.  With exceptions, nearly all accesses to objects managed by
the TM system should use this type so that the CPU will translate the
reference to the right version of the object.

@@

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of the nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

\cfbolz{maybe it's worth explicitly pointing out here that this is one of the
    few places where the \%gs translation must not be used}

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}.
\cfbolz{the details about why the JIT confuses everything should go into
Section~\ref{sec:performance-bench}. here just write "because the integration
with the JIT is incomplete"}
We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT~\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
benchmark.} The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

\subsection{Overhead Breakdown}

\remi{do it on a non-jit build (see reason above)}
\remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}
\remi{update numbers in pypy/TODO}

\begin{itemize}
\item time taken by read \& write barriers
\item time spent committing \& aborting (maybe with different numbers
  of threads; maybe split conflict detection and obj sync on commit)
\item time in GC
\end{itemize}

@@

\subsection{Performance Benchmarks\label{sec:performance-bench}}

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in Figure~\ref{fig:performance-nojit}.

@@

\cfbolz{I suggest making the rest a new section}
\paragraph{JIT benchmarks:} The speedups from enabling the JIT in
these benchmarks range from $10-50\times$. This is why we had to do
without Jython and CPython here, since they would be much further up
in the plots. Also, in order to get more stable results, we increased
the input size of all benchmarks to get reasonable execution times.

@@

\bibitem{stmupdate13} Armin Rigo, Remigius Meier. Update on
  STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}

>>>>

We focus on the \emph{threading} approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} are implementations of this. It requires
great care in order to avoid deadlocks, which is why we follow the TM
approach that provides a \emph{direct} replacement for the GIL. It
does not require careful placing of locks in the right spots.

@@

The x86 CPUs provide a feature which is precisely called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment register
$\%gs$, which is available to applications, to point to the
thread's segment start address.\footnote{The other segment register
$\%fs$ is typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for a
similar purpose: thread-local data -- but a lot of it.}  The translation
from a $SO$ to a LA is done for us by using the ``$\%gs\colon$'' prefix
before any CPU instruction that reads or writes memory using a $SO$
address.  This is very efficient and some compilers support it natively
(e.g.\ clang).

@@

Conflicts can be resolved by aborting one of the two transactions, or in
some cases by pausing one of them until the other commits.  If a commit
succeeds, we copy the changes to modified objects so that they become
visible in other segments.  Note that this only includes objects that
existed before the transaction and have been modified by it.  It does
not include all the new objects created by the current transaction
(there is typically a very high rate of object churn in dynamic languages).
Our garbage collector allocates new objects in pages that are shared in
the first place, so that they do not need to be copied -- and moreover,
such objects will commonly never be modified any more.

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_read(object_t *obj)
void stm_write(object_t *obj)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(obj)
STM_POP_ROOT(obj)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was seen as garbage. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

The type \lstinline!object_t! is special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system should use this type so
that the CPU will translate the reference to the right version of the
object.

@@

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of a thread's nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

\cfbolz{maybe it's worth explicitly pointing out here that this is one of the
    few places where the \%gs translation must not be used}
\arigo{wrong, it is used.  maybe describe a bit more why}

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}.
\cfbolz{the details about why the JIT confuses everything should go into
Section~\ref{sec:performance-bench}. here just write "because the integration
with the JIT is incomplete"}
We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT~\cite{cfbolz09} bases its decision what to
compile on runtime profiling, running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
benchmark.} The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. Note,
it does not measure the memory required for the segment-private areas
like the nursery and read markers. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

%%% TODO: if time permits
% \subsection{Overhead Breakdown}

% \remi{do it on a non-jit build (see reason above)}
% \remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}
% \remi{update numbers in pypy/TODO}

% \begin{itemize}
% \item time taken by read \& write barriers
% \item time spent committing \& aborting (maybe with different numbers
%   of threads; maybe split conflict detection and obj sync on commit)
% \item time in GC
% \end{itemize}

@@

\subsection{Small-Scale Benchmarks\label{sec:performance-bench}}

@@

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in Figure~\ref{fig:performance-nojit}.

@@

\subsection{JIT Benchmarks}
The speedups from enabling the JIT in
these benchmarks range from $10-50\times$. This is why we had to do
without Jython and CPython here, since they would be much further up
in the plots. Also, in order to get more stable results, we increased
the input size of all benchmarks to get reasonable execution times.

@@

\bibitem{stmupdate13} Armin Rigo, Remigius Meier. 2013. Update on
  STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}
